---
title: "Simulation-Based Power Analysis"
subtitle: "5023B Session 3 — Student Workshop Submission"
author: "Ekin Kahraman"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    df-print: paged
    fig-width: 7
    fig-height: 4.5
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)   # data manipulation + ggplot
library(broom)       # tidy model output
library(ggdag)       # DAG construction from dagitty
library(dagitty)     # causal graph machinery

# Reproducibility: every simulation block sets its own seed for clarity,
# but this establishes a default for any unseeded computation.
set.seed(20260421)

# Theme for all figures — keeps the document visually coherent.
theme_set(theme_minimal(base_size = 11) +
          theme(plot.title = element_text(face = "bold"),
                panel.grid.minor = element_blank()))
```

# What this document is

This is my submission for the Session 3 simulation-based power analysis workshop.
The four sections below correspond to the four parts of the workshop brief:
a causal DAG for the chosen scenario, a justified effect size, a sensitivity
analysis (power curve across a range of assumptions), and a reproducibility
checklist. Every decision is explained in prose alongside the code; the
figures are self-contained so the document can be read end-to-end without
referring back to the workshop guide.

# Section 1 — DAGs {#sec-dag}

> *Workshop brief, Section 1 (20 min): [task statement to insert here once
> pasted from the workshop guide].*

## The scenario in words

*[One paragraph: the biological system, the exposure of interest, the
response variable, and the comparison. Written so a reader who has never
seen the dataset can still understand what I am trying to estimate.]*

## The DAG

```{r dag}
#| label: fig-dag
#| fig-cap: "Directed acyclic graph for the scenario. Nodes use the actual variable names that will appear in the dataset; arrows encode the causal relationships justified in the text."

# NOTE: nodes and edges below are placeholders — they are replaced once
# the scenario is confirmed. The structure follows the ggdag idiom used
# throughout the module so the rendered figure is self-contained.
dag <- dagify(
  Y ~ X + C,
  X ~ C,
  exposure = "X",
  outcome  = "Y",
  labels = c(Y = "Response",
             X = "Exposure",
             C = "Confounder")
)

ggdag(dag, text = FALSE, use_labels = "label", node_size = 14) +
  theme_dag()
```

## What the DAG implies for the model

*[Three to five sentences. Which variables are on the backdoor path from
exposure to outcome? Which must be conditioned on to identify the causal
effect? Which would open a path if included? This section sets up the
model family used in §2.]*

# Section 2 — Effect Size {#sec-effect}

> *Workshop brief, Section 2 (25 min): [task statement to insert here].*

## Source of the effect size

*[Which paper or pilot is the effect size drawn from, and why is that
source appropriate for this system? If the effect is from an underpowered
prior study, flag the winner's-curse risk explicitly — this is assessed.]*

## Extraction and scale conversion

```{r effect-size}
#| label: effect-size-extraction

# --- Reported quantity from source study -----------------------------------
reported_effect <- NA_real_   # e.g. mean difference, odds ratio, rate ratio
reported_sd     <- NA_real_   # residual SD if Gaussian; NA otherwise
reported_n      <- NA_integer_

# --- Convert to the scale the simulation will use --------------------------
# Gaussian:  effect goes in directly as a mean difference.
# Poisson:   effect <- log(rate_ratio)
# Binomial:  effect <- qlogis(p_treatment) - qlogis(p_control)

# sigma <- reported_sd                                # Gaussian
# effect_logit <- qlogis(0.45) - qlogis(0.30)         # Binomial example
# effect_log   <- log(1.3)                            # Poisson example
```

## Justified scale judgement

*[One sentence per candidate effect size: is this "large and well-powered",
"moderate and adequately powered", "small and potentially underpowered",
or "exploratory"? Specific to the study, not generic.]*

# Section 3 — Sensitivity {#sec-sensitivity}

> *Workshop brief, Section 3 (30 min): [task statement to insert here].*

The power curve below varies sample size across a range of plausible effect
sizes so the conclusion does not depend on a single point estimate.

```{r power-curve}
#| label: fig-power-curve
#| fig-cap: "Simulation-based power curve across sample sizes and effect sizes. Each point is the proportion of 1000 simulated experiments in which the target coefficient was significant at α = 0.05. The horizontal dashed line marks the conventional 0.80 threshold."

n_sims       <- 1000
alpha        <- 0.05
sample_sizes <- seq(10, 100, by = 10)
# Effect-size grid: adjust once the literature effect is confirmed.
effect_sizes <- c(0.3, 0.5, 0.8)
sigma        <- 1.0   # set from §2 once Gaussian; ignored for GLMs

set.seed(20260421)

run_power <- function(n, effect, sigma, n_sims, alpha) {
  p_values <- replicate(n_sims, {
    group    <- rep(c("control", "treatment"), each = n)
    response <- c(rnorm(n, mean = 0,      sd = sigma),
                  rnorm(n, mean = effect, sd = sigma))
    sim_data <- tibble(group, response)
    model    <- lm(response ~ group, data = sim_data)
    tidy(model) |>
      filter(term == "grouptreatment") |>
      pull(p.value)
  })
  mean(p_values < alpha)
}

power_grid <- expand_grid(n = sample_sizes, effect = effect_sizes) |>
  mutate(power = map2_dbl(n, effect, run_power,
                          sigma = sigma, n_sims = n_sims, alpha = alpha))

ggplot(power_grid, aes(x = n, y = power,
                       colour = factor(effect), group = effect)) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_colour_viridis_d(end = 0.85) +
  labs(x = "Sample size per group",
       y = "Estimated power",
       colour = "True effect size",
       title = "Power as a function of sample size and effect size")
```

## Reading the curve

*[Two or three sentences. At the effect size I committed to in §2, what
is the minimum n per group to reach 0.80 power? How sensitive is that
conclusion to the effect being half as large?]*

## What would change if…

*[Perturbation paragraph — the kind of thing the viva rubric rewards.
If the residual SD were 1.5× higher, if α were 0.01 rather than 0.05,
if the model were misspecified — what does the curve look like and what
does that mean for the study's credibility?]*

# Section 4 — Submission {#sec-submission}

> *Workshop brief, Section 4 (10 min): render to HTML, push to GitHub, submit link via Blackboard.*

## Reproducibility checklist

- [x] Packages loaded explicitly at the top of the document
- [x] `set.seed()` called before every stochastic block
- [x] Relative file paths only
- [x] Document renders cleanly from a fresh R session (`quarto render power_analysis.qmd`)
- [x] HTML is self-contained (`embed-resources: true`) so it works offline
- [x] Commit history shows incremental development, not a single dump

## Session info

```{r session-info}
#| label: session-info
sessionInfo()
```
