---
title: "Simulation-Based Power Analysis"
subtitle: "5023B Session 3 Assessment"
author: "Ekin Kahraman"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    df-print: paged
    fig-width: 7
    fig-height: 4.5
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(broom)
library(ggdag)
library(dagitty)

set.seed(20260421)

theme_set(theme_minimal(base_size = 11) +
          theme(plot.title = element_text(face = "bold"),
                panel.grid.minor = element_blank()))
```

# Section 2 — Building the Study DAG and Choosing an Effect Size

## Part A — The Study DAG

The study is a randomised feeding trial in great tits: nest boxes are randomly
assigned to supplementary feeding or control, and juvenile body mass (g) is
measured at six weeks. Tarsus length (mm) is recorded as a pre-existing
structural measure of body size. The planned model is
`lm(body_mass ~ treatment + tarsus_length)`.

```{r dag-greattit}
#| label: fig-dag-greattit
#| fig-cap: "DAG for the great tit supplementary feeding study. Treatment is randomised, so no arrows point into it. Tarsus length is a structural trait fixed before treatment; it influences body mass but cannot confound the treatment–body mass relationship."

dag_greattit <- dagify(
  body_mass ~ treatment + tarsus_length,
  # Nothing in the study design causes tarsus_length — it is a structural
  # trait determined by genetics and pre-treatment development, so no arrows
  # point into it.
  # Nothing causes treatment — it is randomly assigned to nest boxes,
  # so no arrows point into it either.

  exposure = "treatment",
  outcome  = "body_mass",

  labels = c(
    body_mass     = "Body Mass (g)",
    treatment     = "Supplementary\nFeeding",
    tarsus_length = "Tarsus\nLength (mm)"
  )
)

ggdag(dag_greattit, text = FALSE, use_labels = "label") +
  theme_dag() +
  labs(title = "Great tit supplementary feeding study")
```

```{r dag-adjustment}
#| label: dag-adjustment
ggdag_adjustment_set(dag_greattit)
```

**Discussion response.**
Tarsus length cannot confound the treatment–body mass relationship because
treatment is randomly assigned to nest boxes: randomisation breaks any path
from tarsus length to treatment, and the DAG correctly contains no arrow
from tarsus into treatment. `ggdag_adjustment_set()` confirms the empty
adjustment set — no variable is required to identify the causal effect.
Tarsus length is included in the model as a **precision covariate**: it
explains a substantial portion of the variance in body mass, and
conditioning on it reduces the residual standard deviation of the treatment
coefficient, which increases power without changing the estimand.

## Part B — Where Does Your Effect Size Come From?

For each option I record a single-sentence judgement on whether I would use
it, and a single sentence on the specific evidence that would make me
abandon it.

### Option 1 — Expert Opinion

> `chosen_effect <- 0.75` &nbsp; | &nbsp; `chosen_sigma <- 2.50`

- **Would I use it?** No — a conference remark that cites no primary data
  and reports no standard deviation gives me no way to defend the chosen
  value as anything other than a guess, and a guess is not a defensible
  input to a power analysis.
- **What would make me abandon it?** I would abandon this option unless the
  reviewer's claim were backed by the underlying meta-analysis with
  species-specific effect sizes and pooled residual variances published and
  re-examinable — without those numbers the range "half a gram to one gram"
  is not quantitative evidence.

### Option 2 — A Single Published Study

> `chosen_effect <- 1.80` &nbsp; | &nbsp; `chosen_sigma <- NA` (fallback 2.20)

- **Would I use it?** No — n = 12 per group with a reported p-value of
  0.038 is the textbook winner's-curse / Type M scenario from Gelman &
  Carlin (2014), so the 1.80 g estimate is almost certainly inflated and
  powering from it will understate the sample size I actually need.
- **What would make me abandon it?** I would abandon this option unless the
  raw data were available to re-estimate the effect and its uncertainty
  directly, or a pre-registered replication in great tits (not blue tits)
  reported a comparable magnitude at a larger sample size.

### Option 3 — A Minimum Meaningful Difference

> `chosen_effect <- 0.80` &nbsp; | &nbsp; `chosen_sigma <- 2.10`

- **Would I use it?** Yes — 0.8 g is the smallest effect with a documented
  demographic consequence (overwinter survival and first-year recruitment)
  in the same species at a long-term study site, and the residual SD comes
  from 15 years of tarsus-adjusted Wytham data, so both parameters are
  grounded in the system I am actually studying.
- **What would make me abandon it?** I would abandon this option if the
  Wytham steering committee revised the 0.8 g threshold downward based on
  updated survival models, or if a survival analysis at my own study site
  showed the body-mass–recruitment relationship did not hold there.

### Option 4 — A Well-Powered Replication Study

> `chosen_effect <- 1.50` &nbsp; | &nbsp; `chosen_sigma <- 2.20`

- **Would I use it?** Partly — it is the strongest single evidential source
  (pre-registered, eight sites, n = 400 per treatment, tight CI), but
  powering at the 1.50 g point estimate assumes my study will reproduce the
  central estimate rather than the 1.15 g lower CI bound, which I cannot
  guarantee for a single-site replication.
- **What would make me abandon it?** I would abandon this option if my
  study site were materially different from the eight replication sites
  (e.g. coniferous rather than mixed woodland), or if the lower CI bound
  (1.15 g) were below the minimum biologically meaningful difference
  adopted in Option 3.

### Commitment

I commit to **Option 3**. The 0.8 g threshold is species- and site-matched,
is justified by documented demographic consequences rather than by a single
underpowered estimate, and powering at the minimum biologically meaningful
difference is the conservative choice — if the true treatment effect is
larger than 0.8 g my study will be over-powered (a recoverable outcome),
whereas powering at a larger effect I cannot guarantee would leave me
under-powered to detect biologically meaningful differences.

```{r committed-parameters}
#| label: committed-parameters
chosen_effect <- 0.80   # g; minimum meaningful difference, Option 3
chosen_sigma  <- 2.10   # g; residual SD after tarsus adjustment, Wytham 15-yr
alpha         <- 0.05
n_sims        <- 1000
```

# Section 3 — Power Curves and Sensitivity Analysis

All simulations fit `lm(body_mass ~ treatment + tarsus_length)` on data
generated with a random tarsus covariate. The treatment coefficient is the
quantity of interest; its p-value is extracted and power is the proportion
of 1000 simulated experiments in which that p-value falls below 0.05.

```{r simulation-core}
#| label: simulation-core

# Simulate one experiment and return the p-value for the treatment
# coefficient from lm(body_mass ~ treatment + tarsus_length).
simulate_one <- function(n, effect, sigma) {
  treatment <- rep(c(0, 1), each = n)
  tarsus    <- rnorm(2 * n, mean = 20, sd = 1)
  body_mass <- 10 + 0.5 * tarsus + effect * treatment +
               rnorm(2 * n, mean = 0, sd = sigma)
  model <- lm(body_mass ~ treatment + tarsus)
  tidy(model) |>
    filter(term == "treatment") |>
    pull(p.value)
}

# Estimate power at a single (n, effect, sigma) combination.
estimate_power <- function(n, effect, sigma,
                           n_sims = 1000, alpha = 0.05) {
  p_values <- replicate(n_sims, simulate_one(n, effect, sigma))
  mean(p_values < alpha)
}
```

## Exercise 1 — Effect Size Sensitivity

```{r exercise-1}
#| label: exercise-1

set.seed(20260421)

sample_sizes <- seq(20, 300, by = 20)
effect_grid  <- c(chosen_effect, chosen_effect * 0.7)

ex1 <- expand_grid(n = sample_sizes, effect = effect_grid) |>
  mutate(sigma = chosen_sigma,
         power = pmap_dbl(list(n, effect, sigma), estimate_power,
                          n_sims = n_sims, alpha = alpha))
```

```{r fig-ex1}
#| label: fig-ex1
#| fig-cap: "Power as a function of sample size per treatment group, at the committed effect size (0.80 g) and at a 30% reduction (0.56 g). The dashed line marks the 0.80 power threshold."

ggplot(ex1, aes(x = n, y = power,
                colour = factor(effect, labels = c("0.56 g (−30%)",
                                                   "0.80 g (committed)")),
                group = effect)) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_colour_manual(values = c("0.80 g (committed)" = "#1F78B4",
                                 "0.56 g (−30%)"     = "#E31A1C")) +
  labs(x = "Sample size per treatment group",
       y = "Estimated power",
       colour = "Effect size",
       title = "Exercise 1: effect-size sensitivity")
```

```{r ex1-readoffs}
#| label: ex1-readoffs

min_n_primary <- ex1 |>
  filter(effect == chosen_effect, power >= 0.80) |>
  slice_min(n, n = 1, with_ties = FALSE) |>
  pull(n)

min_n_reduced <- ex1 |>
  filter(effect < chosen_effect, power >= 0.80) |>
  slice_min(n, n = 1, with_ties = FALSE) |>
  pull(n)
```

**Written response.** At the committed effect (0.80 g), the curve first
crosses 0.80 power at **n ≈ `r min_n_primary` birds per treatment group**.
If the true effect is 30% smaller (0.56 g), the same threshold is not
reached until **n ≈ `r min_n_reduced` per group** — roughly double the
committed recommendation. This demonstrates that the recommendation is
asymmetric in its consequences: over-estimating the effect by 30% leads to
a materially underpowered study, while under-estimating it by the same
margin leaves the study over-powered but still correct.

## Exercise 2 — Variance Sensitivity

```{r exercise-2}
#| label: exercise-2

set.seed(20260421)

sigma_grid <- c(chosen_sigma, chosen_sigma * 1.5)

ex2 <- expand_grid(n = sample_sizes, sigma = sigma_grid) |>
  mutate(effect = chosen_effect,
         power  = pmap_dbl(list(n, effect, sigma), estimate_power,
                           n_sims = n_sims, alpha = alpha))
```

```{r fig-ex2}
#| label: fig-ex2
#| fig-cap: "Power as a function of sample size at the committed residual SD (2.10 g) and at a 50% increase (3.15 g). The committed effect (0.80 g) is held constant."

ggplot(ex2, aes(x = n, y = power,
                colour = factor(sigma, labels = c("2.10 g (committed)",
                                                  "3.15 g (+50%)")),
                group = sigma)) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_colour_manual(values = c("2.10 g (committed)" = "#1F78B4",
                                 "3.15 g (+50%)"      = "#33A02C")) +
  labs(x = "Sample size per treatment group",
       y = "Estimated power",
       colour = "Residual SD",
       title = "Exercise 2: variance sensitivity")
```

```{r ex2-readoffs}
#| label: ex2-readoffs

min_n_sigma_hi <- ex2 |>
  filter(sigma > chosen_sigma, power >= 0.80) |>
  slice_min(n, n = 1, with_ties = FALSE) |>
  pull(n)
```

**Written response.** Inflating the residual SD from 2.10 g to 3.15 g
shifts the 0.80-power sample size from **n ≈ `r min_n_primary`** to
**n ≈ `r min_n_sigma_hi` per group**. The shift is of a similar order of
magnitude to the effect-size shift in Exercise 1, so the two assumptions
matter quantitatively as much as each other. However, my residual SD is
anchored in 15 years of Wytham data and is therefore better evidenced than
the effect size, which is a policy threshold rather than an empirical
estimate — meaning the variance axis is less plausible to move than the
effect-size axis.

## Exercise 3 — Sensitivity Grid

```{r exercise-3}
#| label: exercise-3

set.seed(20260421)

ex3 <- expand_grid(
  effect = c(chosen_effect, chosen_effect * 0.7),
  sigma  = c(chosen_sigma, chosen_sigma * 1.5)
) |>
  mutate(n = 30,
         power = pmap_dbl(list(n, effect, sigma), estimate_power,
                          n_sims = n_sims, alpha = alpha))

ex3 |>
  mutate(effect = sprintf("%.2f g", effect),
         sigma  = sprintf("%.2f g", sigma),
         power  = sprintf("%.3f",  power)) |>
  rename(`Effect size` = effect,
         `Residual SD` = sigma,
         `n per group` = n,
         `Estimated power` = power)
```

**Written response.** At the "convenience" sample size of n = 30 per group,
power is well below 0.80 in every cell of the grid. Even under the most
favourable combination (committed effect, committed SD), the study detects
a biologically meaningful difference less than a third of the time. Under
the least favourable combination — a 30% smaller effect and a 50% larger
residual SD — power collapses close to the nominal Type I error rate. A
study with n = 30 per group is therefore not informative about Option 3's
minimum meaningful difference: a non-significant result from such a study
cannot be interpreted as evidence of absence.

# Section 4 — Interpretation and Submission

## Response 1 — What the Curve Recommends

Based on the committed parameters (treatment effect = 0.80 g, residual
SD = 2.10 g after adjusting for tarsus length, α = 0.05), the power curve
first reaches 0.80 at **n ≈ `r min_n_primary` juvenile birds per treatment
group**, i.e. `r 2 * min_n_primary` birds in total. This recommendation
rests on three assumptions made in Section 2: (i) the effect size is the
minimum biologically meaningful difference adopted by the Wytham steering
committee, not the larger point estimates reported in Options 2 and 4;
(ii) the residual SD is the 15-year site-matched Wytham value, not a
between-study pooled estimate; and (iii) tarsus length enters the model as
a precision covariate under randomised treatment assignment, which is
justified by the DAG in Section 2A.

## Response 2 — How Fragile Is This Estimate?

The single assumption my recommendation is most sensitive to is the
**effect size**. A 30% reduction (0.80 → 0.56 g) shifts the required
sample size from **n ≈ `r min_n_primary`** to **n ≈ `r min_n_reduced`** per
group — close to a doubling. The variance assumption produces a shift of
comparable magnitude, but the residual SD is anchored in 15 years of
Wytham data and a reviewer has little reason to move it. The effect size,
by contrast, is a policy threshold about what counts as biologically
meaningful; it is less empirical and therefore more contestable. A
reviewer should treat the sample-size recommendation as reliable
*conditional on* the 0.8 g threshold being the right criterion — if that
threshold is revised, the power analysis must be redone.

## Response 3 — An Honest Methods Statement

> Statistical power was estimated by simulation. For each candidate sample
> size, 1000 datasets were generated under the model
> `body_mass ~ treatment + tarsus_length` with a treatment effect of
> 0.80 g and a residual standard deviation of 2.10 g. The 0.80 g effect is
> the minimum biologically meaningful difference adopted by the Wytham
> Woods long-term study; the residual SD was estimated from 15 years of
> tarsus-adjusted body mass at the same site. Power was defined as the
> proportion of simulations yielding a significant treatment coefficient
> at α = 0.05. A sample of **`r min_n_primary` birds per treatment group**
> achieves 0.80 power. The principal limitation is sensitivity to the
> effect-size assumption: a 30% smaller true effect roughly doubles the
> required sample.

```{r word-count}
#| label: word-count
#| include: false
methods_text <- "Statistical power was estimated by simulation. For each candidate sample size, 1000 datasets were generated under the model body_mass ~ treatment + tarsus_length with a treatment effect of 0.80 g and a residual standard deviation of 2.10 g. The 0.80 g effect is the minimum biologically meaningful difference adopted by the Wytham Woods long-term study; the residual SD was estimated from 15 years of tarsus-adjusted body mass at the same site. Power was defined as the proportion of simulations yielding a significant treatment coefficient at alpha = 0.05. A sample of N birds per treatment group achieves 0.80 power. The principal limitation is sensitivity to the effect-size assumption: a 30% smaller true effect roughly doubles the required sample."
n_words <- length(strsplit(methods_text, "\\s+")[[1]])
```

*Word count: `r n_words` (target 80–120).*

# Submission Checklist

- [x] Author name in YAML header
- [x] `chosen_effect` set to a numeric value matching the Section 2 commitment (0.80)
- [x] `chosen_sigma` set to a numeric value matching the Section 2 commitment (2.10)
- [x] All written response prompts completed
- [x] Both sensitivity plots render without error
- [x] Sensitivity grid table contains four rows
- [x] Final power curve renders without error
- [x] Document renders to HTML without errors
- [ ] GitHub Classroom DAG document (Section 1) committed and pushed — *pending classroom repo access*

## Session info

```{r session-info}
#| label: session-info
sessionInfo()
```
